### 2020-05-16 paper.tex
Citations and Footnotes

Salvaneschi et al. \cite{comprehendingReactiveApplications} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information. It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.

@@

Various ways of implementing RAs emerged over time. The Observer design pattern \cite{observerDesignPattern}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{ownPaper}. Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{SurveyOnReactiveProgramming} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}. The declarativeness of FP finally lead to what we understand under ``Reactive Programming'' today.

@@

An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{REScala} \cite{otherRPDSL}. At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.

@@

Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the greater model of the dependency graph. In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph. Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{REDebuggerSalvaneschi} \cite{RxViz} \cite{RxFiddle} exploring this field. They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.

In matters of time, traditional as well as reactive debuggers work with the present state of the programs execution only. They lack knowledge about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system. An \textbf{omniscient debugger} \cite{omniscientDebuggerPaper} \cite{mozillaRR} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and fore through the program execution trace without the hassle of reconstructing a given failure situation over and over again. This concept comes with its own set of challenges, i.e. demands on storage capacity or efficient user experience \emph{(UX)} design. \cite{omniscientDebuggerProblems}

@@

In order to prepare to give an answer on the first research question, we started with a series of informal interviews. We talked to professionals about their experiestack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage. Eventually, we interviewed five engince in working with RP in their daily jobs. We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{todo}} and ngrx\footnote{\url{todo}} to develop frontend applications. Two of them stated to have more explicit experience with RxJS further. The fifth subject works currently with a combination of akka\footnote{\url{todo}} and akka-streams\footnote{\url{todo}} in Scala to create backend applications. The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.

@@

None of the subjects remembered debugging RP code to be a positive experience, independent of the technology used. All of the subjects use traditional debugging tools. Redux DevTools \footnote{\url{todo}} is in use by all four subjects working with ngrx. There was mention of any more specific debugging tool in regards of akka and akka-streams.

One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented. (ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{todo}} operator to log values. A less invasive way would be preferable as stated.

@@

We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex. Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{url{todo}}. This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe. In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjs-lifecycle} \cite{rxjs-spy-events}. As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels. Without it, they were blind to what actually happens within the system.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Beside the creation of REScala in 2014 \cite{REScala}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{positiveEffectsRP}, Salvaneschi et al. elaborate on the necessity of RP debugging tools which can understand the specific language abstractions. With the plugin \emph{Reactive Inspector} \footnote{\url{http://guidosalva.github.io/reactive-inspector/}} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala. One of its main features is the visualization of the internal dependency graph. The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph. The execution of the program can be interrupted e.g. once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.

@@

\emph{rxfiddle} proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{rxfiddle}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{storyflow} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time. They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.

@@

rxspy-js

\begin{itemize}
	\item Closer look to past Research
	\item \dots and the tools which were built based on it
	\item Set up war story extract to be replicated with existing tooling
\end{itemize}

\begin{itemize}
	\item Brief overview on specific tooling
	\item For frontend, but also other tiers
	\item Rxviz \footnote{\url{https://rxviz.com/}} see also \url{https://rxviz.com/feedback} for input?
	\item rxjs-spy \footnote{\url{https://github.com/cartant/rxjs-spy}}
	\item rxfiddle \cite{10.1145/3180155.3180156}
	\item reactive-inspector \cite{10.1145/2884781.2884815}
	\item TOD \cite{5287015}
\end{itemize}

>>>>

Salvaneschi et al. \cite{7827078} define the Reactive Application \emph{(RA)} as a computer program which processes a continuos, indefinite and ever-changing flow of information. It is contrasted by a traditional application which consumes and processes one piece of input while producing a one-off output result: Think of an image format conversion program which takes an HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} image file for input, processes it and outputs an JPEG file likewise.

@@

Various ways of implementing RAs emerged over time. The Observer design pattern \cite{gamma1995design}, originated in Object Oriented Programming \emph{(OOP)}, is a basic pattern we can find in many of todays API implementations \cite{alabor:2019:reactiveappllications}. Functional Programming \emph{(FP)} shines with specific reactive language extensions like Fran (Haskell) \cite{10.1145/2501654.2501666} or more recently REScala (Scala) \cite{10.1145/2577080.2577083}. The declarativeness of FP finally lead to what we understand under ``Reactive Programming'' today.

@@

An RP environment provides (i) language constructs to describe mentioned graphs, often in form of a domain specific language \emph{(DSL)} \cite{10.1145/2577080.2577083}. At program runtime, a (ii) runtime system takes care that changed values are propagated through the graph, transformations are recomputed where necessary and a consistent state is maintained.

@@

Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the greater model of the dependency graph. In fact, RP provides its own challenges to debuggers: So are call stacks spoiled with invocations of RP runtime system internal matter or breakpoints can only be used on the imperative parts of transformations rather than on the data flow within the graph. Various \textbf{reactive debuggers} and visualization tools for specific RP runtimes emerged over the past few years \cite{10.1145/2884781.2884815} \cite{rxviz} \cite{10.1145/3180155.3180156} exploring this field. They have knowledge of the dependency graph and therefore allow improved interaction with RP concepts.

In matters of time, traditional as well as reactive debuggers work with the present state of the programs execution only. They lack knowledge about what happened before or what is going to happen in the future. This shortcoming is tedious, especially when debugging a problem depending from multiple complex circumstances in a system. An \textbf{omniscient debugger} \cite{5287015} \cite{DBLP:journals/corr/OCallahanJFHNP17} does not interact with the executed program directly. Instead, it records runtime telemetry and provides an interface for later inspection. Engineers can ``time travel'' back and fore through the program execution trace without the hassle of reconstructing a given failure situation over and over again. This concept comes with its own set of challenges, i.e. demands on storage capacity or efficient user experience \emph{(UX)} design. \cite{omniscientDebuggerProblems}

@@

In order to prepare to give an answer on the first research question, we started with a series of informal interviews. We talked to professionals about their experiestack defined through our research area since we were interested in as many different perspective on the topic as possible in this early stage. Eventually, we interviewed five engince in working with RP in their daily jobs. We did not limit the selection of our interview partners by the tech neers: Four work or worked more recently with Angular\footnote{\url{https://angular.io/}} and ngrx\footnote{\url{https://ngrx.io/}} to develop frontend applications. Two of them stated to have more explicit experience with RxJS further. The fifth subject works currently with a combination of akka\footnote{\url{https://akka.io/}} and akka-streams in Scala to create backend applications. The agenda available in Appendix~\ref{sec:interviewquestions} was used to maintain focus during the interview sessions.

@@

None of the subjects remembered debugging RP code to be a positive experience, independent of the technology used. All of the subjects use traditional debugging tools. Redux\footnote{\url{https://redux.js.org/}} DevTools \footnote{\url{https://github.com/reduxjs/redux-devtools}} is in use by all four subjects working with ngrx. There was no mention of any more specific debugging tool in regards of akka and akka-streams.

One particular subject with more in-depth knowledge on RxJS described situations where they would have wished for tools which could ``understand'' the actual \mintinline{TypeScript}{Observable} graph better: (i) When an error is thrown within the function passed to an operator, the generated stack trace reflects where the operator called the function rather than the actual location where the function is implemented. (ii) In order to understand how data traverses through the graph, subjects augmented code manually with the \mintinline{TypeScript}{tap}\footnote{\url{https://rxjs.dev/api/operators/tap}} side effect operator to log values. A less manually invasive way would be preferable as stated.

@@

We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex. Three subjects even claimed using specialized logging functionalities provided through rxjs-spy\footnote{\url{https://cartant.github.io/rxjs-spy/}}. This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe. In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjslifecycle}. As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels. Without it, they were blind to what actually happens within the system.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al. elaborate on the necessity of RP debugging tools which can understand the specific language abstractions. With the plugin \emph{Reactive Inspector} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala. One of its main features is the visualization of the internal dependency graph. The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph. The execution of the program can be interrupted e.g. once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.

@@

\emph{rxfiddle} proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time. They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.

@@
